---
title: "Homework (week 8)"
author: "Sofie Ditmer"
date: "29/10/2020"
output: html_document
---

TASK: Use the rvest library to scrape data of your interest. I have chosen to scrape data about Covid19 from this website: https://www.worldometers.info/coronavirus/#countries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading packages
```{r}
library(rvest) # web-scraping
library(dplyr) # data-wrangling
library(tidyr) # data transformation
library(stringr) # string manipulation
library(janitor) # cleaning headers
library(ggplot2) # visualization
```

Scrape data of interest using the read_html() function from the rvest package.
```{r}
url <- "https://www.worldometers.info/coronavirus/#countries"

# scrape the website
url_html <- read_html(url)
```

Extracting the HTML table through the particular tag for the specific table I want to scrape. I found the tag by looking at the website source code (F12) and finding the table's x-path code. Below I load the html table into a list. 
```{r}
whole_table <- url_html %>% 
 html_nodes(xpath = '//*[@id="main_table_countries_today"]') %>%
 html_table()
whole_table
```

Turning the list into a dataframe: taking the HTML table and ulisting it
```{r}
new_table <- do.call(cbind,unlist(whole_table, recursive = FALSE)) 
head(new_table)
```

Turning the table into a data frame so it is easier to work with
```{r}
data <- as.data.frame(new_table)
```

CLeaning
```{r}
# Removing the first column since it did not contain anything
clean_data <- data[,-1] 

# Removing the plus sign from the columns where they appear
clean_data[, 3] <- gsub("\\+", "", clean_data[, 3])
clean_data[, 5] <- gsub("\\+", "", clean_data[, 5]) 
clean_data[, 7] <- gsub("\\+", "", clean_data[, 7]) 
clean_data[, 7] <- gsub("\\+", "", clean_data[, 7]) 

# Replacing commas with nothing
clean_data[, c(2:14)] <- lapply(clean_data[, c(2:14)], gsub, pattern = ",", replacement = "")

# Changing the class of multiple columns to numeric 
for (i in 2:14){
   clean_data[,i] <- as.numeric(clean_data[,i])
 }

```

Visualizations
```{r}
options(scipen=999) # Disabling scientific notation

# Looking at the total number of Covid cases for each continent
subset(clean_data, Continent == c("North America", "South America", "Asia", "Europe", "Africa", "Australia/Oceania")) %>% 
	ggplot(aes(x = Continent, y = TotalCases )) +
	geom_bar(stat = "identity") +
	labs(x = "Continent", y = "Total Number of Covid19 Cases")

# Looking at the number of deaths for each continent
subset(clean_data, Continent == c("North America", "South America", "Asia", "Europe", "Africa", "Australia/Oceania")) %>% 
	ggplot(aes(x = Continent, y = TotalDeaths)) +
	geom_bar(stat = "identity") +
	labs(x = "Continent", y = "Total Number of Covid19 Deaths")

# Looking at the number of tests for each continent
subset(clean_data, Continent == c("North America", "South America", "Asia", "Europe", "Africa", "Australia/Oceania")) %>% 
	ggplot(aes(x = TotalTests, y = Continent)) +
	geom_bar(stat = "identity") +
	labs(x = "Total Number of Tests", y = "Continent")

# Looking at the number of people recovered
subset(clean_data, Continent == c("North America", "South America", "Asia", "Europe", "Africa", "Australia/Oceania")) %>% 
	ggplot(aes(x = Continent, y = TotalRecovered )) +
	geom_bar(stat = "identity") +
	labs(x = "Continent", y = "Total Number of Recovered Patients")

# Zooming in on the U.S. and Europa
USA_Europe <- clean_data %>% 
	filter(`Country,Other` == "USA" | `Country,Other` == "Europe")

# Number of cases
USA_Europe %>% 
	ggplot(aes(x = `Country,Other`, y = TotalCases)) +
	geom_bar(stat = "identity") +
	labs(x = " ", y = "Total Number of Covid19 Cases")
# Pretty interesting that a country has almost the same number of cases as an entire continent...

# Number of deaths
USA_Europe %>% 
	ggplot(aes(x = `Country,Other`, y = TotalDeaths)) +
	geom_bar(stat = "identity") +
	labs(x = " ", y = "Total Number of Deaths")

# Number of recovered patients
USA_Europe %>% 
	ggplot(aes(x = `Country,Other`, y = TotalRecovered)) +
	geom_bar(stat = "identity") +
	labs(x = " ", y = "Total Number of Recovered Patients")
```
